---
title: "nyt_most_popular"
author: "John Cruz"
date: "2023-03-24"
output: 
  html_document: default
  pdf_document: default
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import Libraries

```{r library, warning=FALSE, message=FALSE}
library(tidyverse)
library(glue)
library(jsonlite)
library(lubridate)
```

## Connect to New York Times API

Using the *httr* library, I will connect to the New York Times API to pull information on the "Most Popular" articles on the [NY Times](http://www.nytimes.com) website based on views. To keep my API key private I incorporate R Studio API *askForPassword* function.  

n6mRruxy9y1R2eNF1PxBkGnzqZbtd4UG
{rstudioapi::askForPassword('Enter NY Times API Key')}

```{r connect-nyt-api}

year <- 2022
month <- 12
api_cnxn <- fromJSON(glue("https://api.nytimes.com/svc/archive/v1/{year}/{month}.json?api-key=n6mRruxy9y1R2eNF1PxBkGnzqZbtd4UG"), flatten = TRUE)
```

## Transform JSON data into Data Frame

Description

```{r json-df}
yearly_df <- 
  as.data.frame(api_cnxn) |> 
  janitor::clean_names() 
```

## Datetime and Columns of Interest

Description

```{r rename}
yearly_df_unnest <- 
  yearly_df |>
  select(response_docs_pub_date, response_docs_section_name, response_docs_keywords) |> 
  unnest(response_docs_keywords, keep_empty = TRUE, names_sep = '_') |> 
  mutate(response_docs_pub_date = str_extract(response_docs_pub_date, "[:graph:]*(?=\\+)"))

yearly_df_unnest$response_docs_pub_date <- 
  yearly_df_unnest$response_docs_pub_date |> 
  ymd_hms()

yearly_df_unnest <-
  yearly_df_unnest |> 
  mutate(hourly_date = floor_date(response_docs_pub_date, unit = 'hour'))

```

## Title

Description

```{r rename}
plot(pressure)
```